---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center"
)
```

# pedmut

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/pedmut)](https://CRAN.R-project.org/package=pedmut)
[![](https://cranlogs.r-pkg.org/badges/grand-total/pedmut?color=yellow)](https://cran.r-project.org/package=pedmut)
[![](https://cranlogs.r-pkg.org/badges/last-month/pedmut?color=yellow)](https://cran.r-project.org/package=pedmut)
<!-- badges: end -->


## Introduction
The **pedmut** package is part of the [**ped suite**](https://magnusdv.github.io/pedsuite/) ecosystem for pedigree analysis in R. Its aim is to provide a framework for modelling mutations in pedigree computations. 

Although **pedmut** is self-contained, its main purpose is to be imported by other **ped suite** packages, like [pedprobr](https://github.com/magnusdv/pedprobr) (marker probabilities and pedigree likelihoods), [forrel](https://github.com/magnusdv/forrel) (forensic pedigree analysis) and [dvir](https://github.com/magnusdv/dvir).

For the theoretical background of mutation models and their properties (stationarity, reversibility, lumpability), I recommend Chapter 5 of [Pedigree analysis in R](https://shop.elsevier.com/books/pedigree-analysis-in-r/vigeland/978-0-12-824430-2), and the references therein.

## Installation
```{r, eval = FALSE}
# The easiest way to get `pedmut` is to install the entire `ped suite`:
install.packages("pedsuite")

# Alternatively, you can install just `pedmut`:
install.packages("pedmut")

# If you need the latest development version, install it from GitHub:
# install.packages("devtools")
devtools::install_github("magnusdv/pedmut")
```

## A simple likelihood example
The examples below require the packages **pedtools** and **pedprobr** in addition to **pedmut**. The first two are core members of the ped suite and can be loaded collectively with `library(pedsuite)`.
```{r, message = F}
library(pedsuite)
library(pedmut)
```

The figure below shows a father and son who are homozygous for different alleles. We assume that the locus is an autosomal marker with two alleles, labelled 1 and 2.

```{r ex1-ped, fig.width=3, fig.height=2.5}
# Create pedigree
x = nuclearPed(father = "fa", mother = "mo", child = "boy")

# Add marker
x = addMarker(x, fa = "1/1", boy = "2/2")

# Plot with genotypes
plot(x, marker = 1)
```

The data clearly constitutes a *Mendelian error*, and gives a likelihood of 0 without mutation modelling:
```{r}
likelihood(x, marker = 1)
```

The following code sets a simple mutation model and recomputes the pedigree likelihood.

```{r}
x2 = setMutationModel(x, marker = 1, model = "equal", rate = 0.1)

likelihood(x2, marker = 1)
```

Under the mutation model, the combination of genotypes is no longer impossible, yielding a non-zero likelihood. To see details about the mutation model, we can use the `mutmod()` accessor: 
```{r}
mutmod(x2, marker = 1)
```


## Mutation models
A mutation matrix in **pedmut** is a stochastic matrix, with each row summing to 1, where the rows and columns are named with allele labels.

Two central functions of package are `mutationMatrix()` and `mutationModel()`. The first constructs a single mutation matrix according to various model specifications. The second produces what is typically required in applications, namely a list of _two_ mutation matrices, named "male" and "female".

The mutation models currently implemented in **pedmut** are:

* `equal`: All mutations equally likely; probability `1-rate` of no mutation. Parameters: `rate`.

* `proportional`: Mutation probabilities are proportional to the target allele frequencies. Parameters: `rate`, `afreq`.

* `random`: This produces a matrix of random numbers, each row normalised to have sum 1. Parameters: `seed`.

* `custom`: Allows any valid mutation matrix to be provided by the user. Parameters: `matrix`.

* `onestep`: Applicable if all alleles are integers. Mutations are allowed only to the nearest integer neighbour. Parameters: `rate`.

* `stepwise`: For this model alleles must be integers or decimal numbers with a single decimal, such as '17.1', indicating a microvariant. Mutation rates depend on whether transitions are within the same group or not, i.e., between integer alleles and microvariants in the latter case. Mutations also depend on the size of the mutation as modelled by the parameter `range`, the relative probability of mutating n+1 steps versus mutating n steps. Parameters: `rate`, `rate2`, `range`.

* `trivial`: Diagonal mutation matrix with 1 on the diagonal. Parameters: None.


## Model properties
Several properties of mutation models are of interest (both theoretical and practical) for likelihood computations. The **pedmut** package provides utility functions for quickly checking these:

* `isStationary(M, afreq)`: Checks if `afreq` is a right eigenvector of the mutation matrix `M`. Stationary models have the desirable property that allele frequencies don't change across generations.

* `isReversible(M, afreq)`: Checks if `M` together with `afreq` form a _reversible_ Markov chain, i.e., that they satisfy the [detailed balance](https://en.wikipedia.org/wiki/Detailed_balance) criterion.

* `isLumpable(M, lump)`: Checks if `M` allows clustering ("lumping") of a given subset of alleles. This implements the necessary and sufficient condition of _strong lumpability_ of Kemeny and Snell (_Finite Markov Chains_, 1976).

* `alwaysLumpable(M)`: Checks if `M` allows lumping of *any* allele subset.

## Further examples

An `equal` model with rate 0.1:
```{r example1}
mutationMatrix("equal", rate = 0.1, alleles = c("a", "b", "c"))
```

Next, a `proportional` model with rate 0.1. Note that this model depends on the allele frequencies.
```{r example2}
mutationMatrix("prop", rate = 0.1, alleles = c("a", "b", "c"), afreq = c(0.7, 0.2, 0.1))
```

To illustrate the `stepwise` model, we recreate the mutation matrix in Section 2.1.3 of Simonsson and Mostad (FSI:Genetics, 2015). This is done as follows:

```{r example3}
mutationMatrix(model = "stepwise", alleles = c("16", "17", "18", "16.1", "17.1"),
               rate = 0.003, rate2 = 0.001, range = 0.5)
```

A simpler version of the `stepwise` model above, is the `onestep` model, in which only the immediate neighbouring integers are reachable by mutation. This model is only applicable when all alleles are integers.

```{r example4}
mutationMatrix(model = "onestep", alleles = c("16", "17", "18"), rate = 0.04)
```
